### 2014-03-01 thesis/src/chapters/chapter-adaptivity.tex
Thesis: problem domain

\subsection{Introduction}
Guaranteeing Quality-of-Service can be seen as continuous monitoring and reacting to undesirable conditions when necessary. In fact, this process is an IT procedure performed by system administrator. Figure \ref{fig:it-process} illustrates an example of such procedure. While system administrator can be directly responsible for manually performing steps depicted in figure \ref{fig:it-process}, it is much more favourable to automatize this process leading to a self-manageable system. Not only does it lead to efficiency of an IT process but also to its effectiveness \cite{IBM06}. It is achieved through:
\begin{itemize}
  \item \emph{rapid process initiation} - components auto-initiate actions based on information derived from a system
  \item \emph{reduced time and skill requirements} - automatization of IT processes makes them easier and less troublesome what is especially important for skill-intensive, error-prone and long lasting tasks
\end{itemize}

@@

What control loop is made of is:
\begin{inparaenum}[1)]
 \item collection of data
 \item data analysis
 \item making a decisions
 \item enforcing actions
\end{inparaenum}
Figure \ref{fig:feedback-loop} depicts a generic feedback loop, giving context associated with each phase. Successive sections details aforementioned phases.

@@

\subsection{Collecting data}
During this phase, relevant data is collected from sensors. After that, data can be aggregated, filtered or stored to provide a reference point during future feedback loop evaluation. During this phase, there are a few concerns that should be investigated \cite{brun2009engineering}:
\begin{itemize}
 \item sample rate
 \item probe reliability
 \item data format
\end{itemize}

Almost any cloud platform providers incorporates into its solution monitoring component, responsible for data collection. Mechanisms used vary from providers to provider and may involve mixture of protocols and standards. For example, Carina \cite{Carina} uses OpenNebula IM and VMM drivers relaying on a plaintext data, transferred through a SSH connection.

\subsection{Analysis}
Analysis part applies mathematical models to reason about data gathered during previous phase. It can leverage variety of approaches and be driven by a customer-defined policies.

As \cite{brun2009engineering} specifies, it is important to take into account following concerns during analysis phase:
\begin{itemize}
 \item applying historical data and existing patterns to current situation
 \item archiving data
 \item applying adequate model to current situation
 \item model's stability
\end{itemize}

@@

As it was already mentioned, models are often controlled by a policies that denotes a condition which, when satisfied, triggers an action, harnessing environment's disturbance. Currently, industry leaders supports \cite{AmazonAutoScaling} two kind of policies:
\begin{itemize}
 \item \textit{expression based} - allows to define how to scale application in response to changing conditions, which include factors such as memory, CPU usage, cost or some indirect, calculated metrics
 \item \textit{scheduled} - allows to scale an application in response to predictable load changes. For example, traffic increases during the weekends and decreases on working days. Hence, that predictable traffic patterns is used to scale application based on current time.
\end{itemize}

@@

\subsection{Decision arrangement}
Before implementing changes in an environment, decision what resources have to be tuned in order to reach a desirable state is made. Decision should be made on the basis of:
\begin{itemize}
 \item risk analysis - analysing how change effects whole environment
 \item prioritising identified problems and resources
\end{itemize}

In a case of platform-as-a-service providers, none of them explicitly specifies how a change plan is composed.

\subsection{Actions}
Finally, a change plan is implemented in a system. Actions can cover a variety of scaling and resource tuning operations. Next chapter investigates problem of scaling applications in detail.

>>>>

\subsection{Introduction}
Guaranteeing Quality-of-Service can be seen as continuous monitoring and reacting to undesirable conditions when necessary. In fact, this process is an IT procedure performed by a system administrator. Figure \ref{fig:it-process} illustrates an example of such procedure. While system administrator can be directly responsible for manually performing steps depicted in figure \ref{fig:it-process}, it is much more favourable to automatize this process leading to a self-manageable system. Not only does it lead to efficiency of an IT process but also to its effectiveness \cite{IBM06}. It is achieved through:
\begin{itemize}
  \item \emph{rapid process initiation} - components auto-initiate actions based on information derived from a system
  \item \emph{reduced time and skill requirements} - automatization of IT processes makes them easier and less troublesome what is especially important for skill-intensive, error-prone and long lasting tasks
\end{itemize}

@@

What control loop is made of is:
\begin{inparaenum}[1)]
 \item collection of data
 \item data analysis
 \item planning decisions
 \item enforcing actions.
\end{inparaenum} Figure \ref{fig:feedback-loop} depicts such a generic feedback loop. Successive sections detail aforementioned phases.

@@

\subsection{Collecting data}
During this phase, relevant data is collected from sensors. After that, data can be aggregated, filtered or stored to provide a reference point for future feedback loop evaluations. During this phase, there are a few matters that should be examined \cite{brun2009engineering}:
\begin{itemize}
 \item sample rate
 \item probe reliability
 \item data format
\end{itemize}

Mechanisms used by a monitoring component vary from provider to provider and may involve a mixture of protocols and standards. For example, Carina \cite{Carina} uses OpenNebula IM and VMM drivers that relays on a plaintext data transferred through a SSH connection.

\subsection{Analysis}
Analysis part applies mathematical models to reason about data gathered during previous phase. It can leverage a variety of approaches and be driven by a customer-defined policies. As \cite{brun2009engineering} specifies, it is vital to take into account following concerns during analysis phase:
\begin{itemize}
 \item applying historical data and existing patterns to a current situation
 \item archiving data
 \item applying adequate model to a current situation
 \item model's stability
\end{itemize}

@@

As it was already mentioned, models are often controlled by policies that denote a condition which, when satisfied, triggers an action harnessing environment's disturbance. Currently, industry leaders supports \cite{AmazonAutoScaling} two kind of policies:
\begin{itemize}
 \item \textit{expression based policy} - it allows to define how to scale application in response to changing conditions, which include factors such as memory, CPU usage, cost or some indirect, calculated metrics
 \item \textit{scheduled policy} - it allows to scale an application in response to predictable load changes. For example, traffic increases during the weekends and decreases on working days. Hence, that predictable traffic patterns is used to scale application based on current time.
\end{itemize}

@@

\subsection{Decision arrangement}
Before implementing changes in an environment, decision what resources have to be tuned in order to reach a desirable state is made. Decision should be made on the basis of:
\begin{itemize}
 \item risk analysis - analysing how a change affects environment
 \item prioritising identified problems and resources
\end{itemize}

\subsection{Actions}
Finally, a change plan is implemented in a system. Actions can cover a variety of scaling and resource tuning operations. Next chapter investigates problem of scaling an application in detail.

### 2014-03-01 thesis/src/chapters/chapter-scalability.tex
Thesis: problem domain

The reason why scaling application lies in our area of interest is the fact that it is widely accepted measure for improving application performance, consequently increasing offered Quality-of-Service. Enriching system with capability to scale entails avoiding additional costs that are related to coping with excessive traffic. In some cases, these costs may be caused by not handling extra traffic at all and may involve aspects such as: increased response time, processing overhead, space, memory, or money \cite{Bo00}.

While scalability is a widely used term, it still lacks a clear and concise definition. Over the time, there were a few attempts to define it, yet not all of them were claimed as successful \cite{Hi90} \cite{DuRoWi06}. Hence, it is necessary to clarify this term before going into further discussion. Instinctively, scalability is perceived as ability of a system to accommodate an increasing number of elements or objects to process. In particular, we can point out different types of scalability that are affected by increased number of requests: \cite{Bo00}:
\begin{itemize}
	\item \textit{load scalability} - ability to work without delays and unproductive resource consumption at light, moderate, or heavy loads while making good use of available resources. Factors that may hinder load scalability include: scheduling shared resource, self-expansion, inadequate exploitation of parallelism
	\item \textit{space scalability} - memory requirements do not grow to intolerable levels as the number of items system supports increases
	\item \textit{space-time scalability} - system continues to function gracefully as the number of objects it encompasses increases by orders of magnitude
	\item \textit{structural scalability} - implementation or standards do not impede the growth of the number of objects system encompasses
\end{itemize}
Although, all of the aforementioned aspects are vital for any application, our work focuses solely on the first type of scalability. The reasoning behind this statement is that, while all of these scalability types lies in direct responsibility of an application developer, the load scalability can be additionally improved by adding additional resources to a system. This brings us to a question what kind of resources are used by an application or more appropriately in context of this dissertation: \textit{what kind of resources can we add to improve application performance?} Required resources varies from an application to an application. However, among the most common ones we can distinguish:
\begin{itemize}
	\item CPU
	\item memory
	\item storage
	\item network bandwidth
\end{itemize}

It is commonly agreed that there are two main possible ways the resource can be added:
\begin{itemize}
	\item \textit{horizontal scaling (scaling out)} - adding more nodes to a system, such as servers in a context of distributed application
	\item \textit{vertical scaling (scaling up)} - increasing capacity of a single node in a system, i.e. adding additional memory, CPU, storage, etc.
\end{itemize}

What makes scaling application particularly interesting are the benefits offered by a cloud computing, especially the illusion of a virtually infinite computing infrastructure \cite{VaRoBu11}. Making use of virtualization technologies, which often underpins cloud computing platform, allows for resource manipulation in a dynamic, on-demand manner. Although, cloud computing offers additional scaling capabilities, it increases solution complexity since they operate in different layers: server, platform, network as stated in \cite{VaRoBu11}. However, since platform containers are often represented either as virtual machines or another isolated environment (e.g. OpenShift leverages SELinux and cgroups) they are similar in nature to server scaling and supports both scaling up and out. Therefore, the remaining of this chapter is focused solely on server scaling, omitting network scaling as it lays outside of scope of this dissertation.

Having that said, common sense dictates that adding resources is only a part of the success - it should be accompanied by tuning application platform configuration. For example, adding supplementary CPUs without increasing thread pool size that handle requests makes a little sense. Similarly, in context of a Java application, we have to increase heap size, to make a good use of extra memory. While importance of application tuning cannot be underestimated, its detailed analysis lies outside of the scope of this dissertation. Figure \ref{fig:scalability-layers} presents different scalability layers and actions that can be taken at each level to improve application performance.

@@

\subsection{Horizontal scaling}
As outlined in previous section, horizontal scaling is about adding supplementary nodes to a system. As it is common to cloud computing, nodes are represented as virtual machines and this assumption is used in further discussion. Consequently, adding server comes down to cloning a new virtual machine from a template and possibly installing additional software and reconfiguring it later. While mechanism of creating new virtual machine from a template is offered literally in every IaaS platform currently available (OpenStack \cite{OpenStack}, OpenNebula \cite{OpenNebula}, CloudStack \cite{CloudStack} or Eucalyptus \cite{Eucalyptus} to name a few) and is similar in manner, the underlying hardware and virtualization mechanism determines how fast provisioning is done.

Provisioning new server is only a first step in scaling an application, it is required to configure load balancing mechanism to make use of additional node. The two important aspects that have to be consider are: load-balancing algorithms and scalability.

\subsubsection{Load-balancing algorithms}
Generally, there are two types of load-balancers: hardware and software based. Due to the dynamic nature of system under consideration, we focus only on the latter as it offers a greater deal of flexibility. Among the most common algorithms we can distinguish \cite{HaProxyDoc}:
\begin{itemize}
 \item \textit{round-robin scheduling} - request are sent to successive nodes, according to their weights. This algorithm is fairest when the server's processing time remains equally distributed \cite{HaProxyDoc}
 \item \textit{least connection} - the server with the lowest number of connections receives the connection
 \item \textit{source routing} - source IP address is hashed, the same client IP address always reaches the same server
 \item \textit{URI hashing} - URI that designates resource is hashed and divided by the total weight of the running servers. Such hash designates which server that receives the request. In practice, this algorithm is commonly used with proxy caches and anti-virus proxies in order to maximize the cache hit rate.
 \item \textit{request counting algorithm} - load is distributed the requests among the various workers, ensuring that each gets their configured share of the number of requests
 \item \textit{weighted traffic counting algorithm} - variation of above-mentioned algorithm with a difference that it is focused on bytes rather than number of request
 \item \textit{pending request counting algorithm} - scheduler keeps track of how many requests each worker is assigned at present. A new request is automatically assigned to the worker with the lowest number of active requests
\end{itemize}

Situation gets further complicated when considering real-world web application that sends user information using cookies, what imposes requirement on load-balancer for session stickiness \cite{StBaMa11}.

\subsubsection{Load-balancing scalability}
Although, it may seem that balancing workloads eliminates problem of a single point of failure (SPOF) among different servers, it is in fact shifted to load-balancing layer. In other words, load-balancer becomes a new SPOF. Therefore, in cases where high availability is required, multi-tiered load balancing architecture should be considered. This, however, seems not to be a case among IaaS or PaaS providers - none of them unequivocally specifies whether their provide redundancy at load-balancer level.

@@

While there are many load-balancers available on the market, following are credited to be most popular:
\begin{itemize}
 \item \textbf{HAProxy} \cite{HAProxy} - load-balancer initially written by Willy Tarreau. Noticeably, it's used by OpenShift \cite{OpenShift} to distribute load among gears \cite{OpenShiftScaling}
 \item \textbf{BIG-IP Local Traffic Manager (LTM)} - solution offered by F5 \cite{F5}. Although LTM is a hardware solution, omitted in this section, it also has also its virtualized counterpart.
 \item \textbf{Apache HTTPD} \cite{ApacheHTTPD} - popular HTTP server. When enhanced with additional modules, it can behave like a proxy or load-balancer. Over the time, there were several attempts to develop such modules: mod\_jk \cite{ApacheModJk}, mod\_proxy\_balancer \cite{ApacheModProxyBalancer}, to name a few. While the former is purely AJP13 oriented, the latter supports different protocols: HTTP, FTP and AJP13. As a consequence, only mod\_proxy\_balancer was taken into account during comparision. 
 \item \textbf{Zeus Load Balancer} \cite{Zeus} - load balancer offered by layer47 \cite{Layer47}, it incorporates techniques such as Layer-7 traffic management and content aware algorithms
 \item \textbf{Zeus Global Load Balancer} \cite{ZeusGlobal} - load balancer offered by layer47 \cite{Layer47}, it is specialised in balancing load among globally distributed data centres

@@

-- optimized HTTP header analysis: headers are parsed an interpreted on the fly

@@

BIG-IP Local Traffic Manager \cite{LTM} &
-- managing at application services level rather than at individual devices and objects

@@

\newpage
\subsection{Vertical scaling}
Essentially, vertical scaling is concentrated upon increasing capacity of single node. Again, when considering technical advancements that comes with cloud computing and virtualization, we can differ two categories of scaling: virtual machine resizing and virtual machines replacement. This distinction is dictated by limitation hypervisors - not all of them are able to resize virtual machine without shutting it down.

@@

\subsubsection{Virtual machine replacement}
As it was highlighted in previous section, reasoning behind virtual machine replacement is that, in case when dynamic resizing is not possible, a new virtual machine with a desired configuration can be provisioned and replace the old one. Since this is a basic operation, all above-mentioned hypervisors supports this scenario as long as required resources are available.

>>>>

Scaling is a widely accepted measure of improving application performance leading to increase of offered Quality-of-Service. Enriching system with capability to scale entails avoiding additional costs that are related to coping with excessive traffic. In some cases, these costs may be caused by not handling extra traffic at all and may involve aspects such as: increased response time, processing overhead, space, memory, or money \cite{Bo00}.

While scalability is a widely used term, it still lacks a clear and concise definition. Over the time, there were a few attempts to define it, yet not all of them were claimed as successful \cite{Hi90} \cite{DuRoWi06}. Hence, it is necessary to clarify this term before going into further discussion. Instinctively, scalability is perceived as an ability of a system to accommodate an increasing number of elements or objects to process. In particular, we can point out different types of scalability that are affected by increased number of requests: \cite{Bo00}:
\begin{itemize}
	\item \textit{load scalability} - an ability to work without delays and unproductive resource consumption at light, moderate, or heavy loads while making good use of available resources. Factors that may hinder load scalability include: scheduling shared resource, self-expansion, inadequate exploitation of parallelism
	\item \textit{space scalability} - memory requirements do not grow to intolerable levels as the number of items system supports increases
	\item \textit{space-time scalability} - system continues to function gracefully as the number of objects it encompasses increases by orders of magnitude
	\item \textit{structural scalability} - implementation or standards do not impede the growth of the number of objects system encompasses
\end{itemize}
Although, all of the aforementioned aspects are vital for any application, our work focuses solely on the first type of scalability. The reasoning behind this is that while all of these scalability types lies in direct responsibility of an application developer, the load scalability can be additionally improved by adding additional resources to a system. This brings us to a question what kind of resources are used by an application or more appropriately in context of this dissertation: \textit{what kind of resources can we add to improve application performance?} Required resources vary from application to application. However, among the most common one we can distinguish:
\begin{itemize}
	\item CPU
	\item memory
	\item storage
	\item network bandwidth
\end{itemize}

It is commonly agreed that there are two ways of adding a resource:
\begin{itemize}
	\item \textit{horizontal scaling (scaling out)} - adding more nodes to a system, such as servers in a context of distributed application
	\item \textit{vertical scaling (scaling up)} - increasing capacity of a single node in a system, i.e. adding additional memory, CPU, storage, etc.
\end{itemize}

Cloud computing makes scaling application especially interesting due to the illusion of a virtually infinite computing infrastructure \cite{VaRoBu11}. Virtualization technologies, which often underpins cloud computing platform, allows for resource manipulation in a dynamic, on-demand manner. Although cloud computing offers additional scaling capabilities, it increases solution complexity because it operates in different layers: server, platform and network as stated in \cite{VaRoBu11}. However, since platform containers are often represented either as virtual machines or another isolated environment (e.g. OpenShift leverages SELinux and cgroups) they are similar to server scaling and supports both scaling up and out. Therefore, the remaining of this chapter is focused solely on server scaling, omitting network scaling as it lays outside of scope of this dissertation.

Adding resources is only a part of the success - it should be accompanied by tuning application platform's configuration. For example, adding supplementary CPUs without increasing thread pool size makes a little sense. Similarly, in context of a Java application, we have to increase heap size, to make a good use of extra memory. While importance of application tuning cannot be underestimated, its detailed analysis lays outside of the scope of this dissertation. Figure \ref{fig:scalability-layers} presents different scalability layers and actions that can be taken at each level to improve application performance.

@@

\subsection{Horizontal scaling}
As outlined in previous section, horizontal scaling is about adding supplementary nodes to a system. In cloud computing, it is common to represent nodes as virtual machines and this assumption is used in further discussion. Consequently, adding server comes down to cloning a new virtual machine from a template and possibly installing additional software and reconfiguring it later. While mechanism of creating new virtual machine from a template is offered literally in every IaaS platform currently available (OpenStack \cite{OpenStack}, OpenNebula \cite{OpenNebula}, CloudStack \cite{CloudStack} or Eucalyptus \cite{Eucalyptus} to name a few) and is similar in manner, the underlying hardware and virtualization mechanism determines how fast provisioning is done.

Provisioning new server is only a first step in scaling an application, it is required to configure load balancing mechanism to make use of additional node. The two important aspects that have to be consider are: load-balancing algorithms and load-balancing scalability.

\subsubsection{Load-balancing algorithms}
Generally, there are two types of load-balancers: hardware and software based. Due to the dynamic nature of application requirements, we focus only on the latter as it offers a greater deal of flexibility. Among the most common algorithms we can distinguish \cite{HaProxyDoc}:
\begin{itemize}
 \item \textit{round-robin scheduling} - request are sent to successive nodes, according to their weights. This algorithm is fairest when processing time remains equally distributed \cite{HaProxyDoc}
 \item \textit{least connection} - the server with the lowest number of connections receives the connection
 \item \textit{source routing} - source IP address is hashed, the same client IP address always reaches the same server
 \item \textit{URI hashing} - URI that designates resource is hashed and divided by the total weight of the running servers. Such hash designates which server receives the request. In practice, this algorithm is commonly used with proxy caches and anti-virus proxies in order to maximize the cache hit rate.
 \item \textit{request counting algorithm} - load is distributed the requests among the various workers, ensuring that each gets its configured share of the number of requests
 \item \textit{weighted traffic counting algorithm} - variation of the above-mentioned algorithm. However, it is focused on bytes rather than number of request
 \item \textit{pending request counting algorithm} - scheduler keeps track of how many requests is assigned to each worker. A new request is automatically assigned to the worker with the lowest number of active requests
\end{itemize}

Situation gets further complicated when considering real-world web application that sends user information using cookies, what imposes on load-balancer requirement for session stickiness \cite{StBaMa11}.

\subsubsection{Load-balancing scalability}
Although, it may seem that balancing workloads eliminates problem of a single point of failure (SPOF) among different servers, it is in fact shifted to load-balancing layer. In other words, load-balancer becomes a new SPOF. Therefore, in cases where high availability is required, multi-tiered load balancing architecture should be considered. This, however, seems not to be a case among current IaaS or PaaS providers - none of them unequivocally specifies whether their provide redundancy at load-balancer level.

@@

While there are many load-balancers available on the market, following are credited to be the most popular:
\begin{itemize}
 \item \textbf{HAProxy} \cite{HAProxy} - load-balancer initially written by Willy Tarreau. Noticeably, it's used by OpenShift \cite{OpenShift} to distribute load among gears \cite{OpenShiftScaling}
 \item \textbf{BIG-IP Local Traffic Manager (LTM)} - solution offered by F5 \cite{F5}. Although LTM is a hardware solution, omitted in this section, it also has its virtualized counterpart.
 \item \textbf{Apache HTTPD} \cite{ApacheHTTPD} - popular HTTP server. When enhanced with additional modules, it can behave like a proxy or load-balancer. Over the time, there were several attempts to develop such modules: mod\_jk \cite{ApacheModJk}, mod\_proxy\_balancer \cite{ApacheModProxyBalancer}, to name a few. While the former is purely AJP13 oriented, the latter supports different protocols: HTTP, FTP and AJP13. As a consequence, only mod\_proxy\_balancer was taken into account during comparision. 
 \item \textbf{Zeus Load Balancer} \cite{Zeus} - load balancer offered by layer47 \cite{Layer47}, it incorporates techniques such as Layer-7 traffic management and content aware algorithms
 \item \textbf{Zeus Global Load Balancer} \cite{ZeusGlobal} - load balancer offered by layer47 \cite{Layer47}, it is specialised in balancing load among globally distributed data centres

@@

-- optimized HTTP header analysis: headers are parsed and interpreted on the fly

@@

BIG-IP Local Traffic Manager \cite{LTM} &
-- managing application services level rather than individual devices and objects

@@

\newpage
\subsection{Vertical scaling}
Essentially, vertical scaling is concentrated upon increasing capacity of a single node. Again, when considering technical advancements that comes with cloud computing and virtualization, we can differ two categories of scaling: virtual machine resizing and virtual machines replacement. This distinction is dictated by limitation of hypervisors - not all of them are able to resize virtual machine without shutting it down.

@@

\subsubsection{Virtual machine replacement}
As it was highlighted in previous section, reasoning behind virtual machine replacement is that, in case when dynamic resizing is not possible, a new virtual machine with a desired configuration can be provisioned and replace the old one. This is a basic operation and therefore all above-mentioned hypervisors support it as long as required resources are available.
